---
title: "Tests for Randomness"

author: "Ananyo Dey
      <br>
      Kaustav Paul"
institute: Indian Statistical Institute, Delhi Center
Date: "2024.04.11"
format: 
  revealjs: 
    incremental: true
    code-fold: true
    transition: slide
    theme: serif
    fontsize: 150%
execute: 
  echo: true
editor: visual
---

# Some Statistical Tests for Goodness of Fit

## Chi-square Goodness of Fit Test

Let $X_1, X_2, ..., X_n \sim F(.)$ independently.

We consider mutually exclusive and exhaustive classes $C_1, C_2,...,C_k$ such that

$$\mathbb{P}(X \in C_i) = p_i$$ We want to test the hypothesis,

$$H_0:p_i = p_i^0 , \text{for all } i \quad \text{vs} \quad H_1 : p_i \neq p_i^0, \text{for at least one } i$$

Let $f_i$ be the observed frequency of the $i^{th}$ class in the sample. The observed frequency of the $i^{th}$ class is given by $O_i = np_i$ for all $i$.

## Chi-square Goodness of Fit Test

The chi-square goodness of fit statistic is given by,

$$Q = \sum_{i=1}^{k} \frac{(f_i - np_i^0)^2}{np_i^0}$$ Under $H_0$, $Q \xrightarrow{d} \chi^2_{k-1}$ as $n \rightarrow \infty$

**Critical Region:** We reject $H_0$ for large values of observed $Q$. For large $n$ we reject $H_0$ if $Q_{obs} > \chi^2_{k-1}(\alpha)$, where $\chi^2_{k-1}(\alpha)$ is the upper $\alpha$ point of a $\chi^2_{k-1}$ distribution.

## Kolmogorov-Smirnov Test

Suppose $X_1, X_2, ..., X_n \sim F(.)$ independently.

Let $F_0(.)$ be a known distribution function. We want to test the hypotheses,

$$H_0 : F(x) = F_0(x), \text{for all  } x \quad vs \quad F(x) \neq F_0(x), \text{for at least one }x$$

Let $S_n(.)$ be the empirical cdf defined as,

$$S_n(x) = \frac{1}{n} \sum _{i = 1} ^n \mathbb{I}\{X_i \leq x\}$$

Then $$nS_n \sim Bin(n,F(x))$$

## Kolmogorov-Smirnov Test

The Kolmogorov - Smirnov Test Statistic is given by

$$D_n = \sup_{x} |S_n(x) - F_0(x)|$$

**Critical Region:** We reject $H_0$ for large values of $Q_{obs}$.

# Empirical Tests

## The Setup

::: incremental
-   Let us consider a sequence of numbers $$U_1, U_2, U_3, ...\quad (1)$$ of real numbers where each $0 < U_i < 1$. We need to test whether it is a random sample from $U(0,1)$.

-   Some of the tests are designed primarily for integer-valued sequences, instead of the real-valued sequence (1). In this case, the auxiliary sequence $$Y_1, Y_2, ... \quad (2)$$ defined by the rule $$ Y_n = \lfloor dU_n \rfloor \quad ...(3)$$ is used instead. So testing the uniformity of $\{U_n\}$ between 0 and 1 is equivalent to testing uniformity of $\{Y_n\}$ in the set $\{0, 1, ..., d-1\}$.
:::

# Frequency Test

## Test Procedure

::: incremental
-   We are provided with a sequence of real numbers $$U_1, U_2,...$$ We need to test $$H_0: U_1, U_2, ... \sim U(0,1) \text{ independently} \quad \text{vs} \quad H_1: \text{not } H_0$$

-   For an integer $d$ we construct the sequence $\{Y_n\}$ as discussed earlier.

-   If $H_0$ is true, $\{Y_n\}$ is a random sample from $\{0, 1, ..., d-1\}$.

-   We count the number of occurences of $r$ in $\{Y_n\}$ for all $0 \leq r \leq d-1$.

-   Under $H_0$, expected frequency of each $r$ in $\{Y_n\}$ is $\frac{n}{d}$.

-   Now we use the $\chi^2$-goodness of fit test over $\{Y_n\}$ using $\{0, 1, ..., d-1\}$ as the $d$ classes.
:::

## Algorithm to Calculate Frequencies

```{r echo = F, warning=F, fig.height=7}
library(DiagrammeR)
graph <- "
digraph flowchart {

  # Node definitions
  A [shape=box, label=\"U_n\", color=\"blue\"];
  B [shape=box, label=\"Y_n\", color=\"blue\"];
  C [shape=box, label=\"count[r] = 0\", color=\"blue\"];
  D [shape=ellipse, label=\"Is r = Y_i?\", color=\"orange\"]
  E [shape=box, label=\"count[r] = count[r] + 1\", color=\"blue\"]
  F [shape=box, label=\"i = i+1\", color=\"blue\"]
  G [shape=ellipse, label=\"Is i <= n?\", color=\"orange\"]
  H [shape = box, label=\"r = r+1\", color=\"blue\"]
  I [shape=ellipse, label=\"Is r<d?\", color=\"orange\"]
  J [shape=box, label=\"End\", color=\"blue\"]
  
  
  
  
  # Edges to indicate the flow
  A -> B [label=\"Divide by d and take floor\"]
  B -> C [label=\"set r=0\"]
  C -> D [label = \"set i=1\"]
  D -> E [label=\"yes\"]
  E -> F 
  D -> F [label=\"no\"]
  F -> G 
  G -> D [label=\"yes\"]
  G -> H [label=\"no\"]
  H -> I 
  I -> C [label=\"yes\"]
  I -> J [label=\"no\"]
  

}

"

grViz(graph)
```


```{r}
freq_test <- function (u, d) 
{
  seq = 0:(d-1)
  integernum <- floor(u * d)
  obsnum <- sapply(seq, function(x) sum(integernum == x))
  expnum <- length(u)/length(seq)
  residu <- (obsnum - expnum)/sqrt(expnum)
  stat <- sum(residu^2)
  pvalue <- pchisq(stat, length(seq) - 1, lower.tail = FALSE)
  options(digits = 2)
  df <- data.frame(1:length(obsnum), obsnum, expnum)
  colnames(df) <- c("Y_j", "Observed Freq", "Expected Freq")
  
  res <- list(statistic = stat, parameter = length(seq) - 1, 
              p.value = pvalue, df)
  return(res)
}
```

## Implementation of the `freq_test` Function

**Under $H_0$:**

$$U_1, U_2, U_3, ... \sim U(0,1)$$

```{r}
sample <- runif(1000)
d <- 50
freq_test(sample, d)
```

